% Exercises (Exercises) for Chapter 18

\section*{Exercises}
\addcontentsline{toc}{section}{Exercises}

\subsection*{Easy}

\begin{problem}[MDP Definition]
Define the components of a Markov Decision Process.

\textbf{Hint:} States, actions, rewards, transition dynamics, discount factor.
\end{problem}

\begin{problem}[Value Function]
Explain the difference between $V(s)$ and $Q(s,a)$.

\textbf{Hint:} State value vs. action-value.
\end{problem}

\begin{problem}[Policy Types]
Contrast deterministic and stochastic policies.

\textbf{Hint:} Mapping vs. distribution over actions.
\end{problem}

\begin{problem}[Exploration vs. Exploitation]
Give two exploration strategies in RL.

\textbf{Hint:} $\epsilon$-greedy; UCB; entropy regularisation.
\end{problem}

\subsection*{Medium}

\begin{problem}[Bellman Equation]
Derive the Bellman equation for $Q(s,a)$.

\textbf{Hint:} Recursive relationship with successor states.
\end{problem}

\begin{problem}[Policy Gradient]
Explain why policy gradient methods are useful for continuous action spaces.

\textbf{Hint:} Direct parameterisation; differentiability.
\end{problem}

\subsection*{Hard}

\begin{problem}[Actor-Critic Derivation]
Derive the advantage actor-critic update rule.

\textbf{Hint:} Baseline subtraction; variance reduction.
\end{problem}

\begin{problem}[Off-Policy Correction]
Analyse importance sampling for off-policy learning and its variance.

\textbf{Hint:} Likelihood ratio; distribution mismatch.
\end{problem}


\begin{problem}[Advanced Topic 1]
Explain a key concept from this chapter and its practical applications.

\textbf{Hint:} Consider the theoretical foundations and real-world implications.
\end{problem}

\begin{problem}[Advanced Topic 2]
Analyse the relationship between different techniques covered in this chapter.

\textbf{Hint:} Look for connections and trade-offs between methods.
\end{problem}

\begin{problem}[Advanced Topic 3]
Design an experiment to test a hypothesis related to this chapter's content.

\textbf{Hint:} Consider experimental design, metrics, and potential confounding factors.
\end{problem}

\begin{problem}[Advanced Topic 4]
Compare different approaches to solving a problem from this chapter.

\textbf{Hint:} Consider computational complexity, accuracy, and practical considerations.
\end{problem}

\begin{problem}[Advanced Topic 5]
Derive a mathematical relationship or prove a theorem from this chapter.

\textbf{Hint:} Start with the definitions and work through the logical steps.
\end{problem}

\begin{problem}[Advanced Topic 6]
Implement a practical solution to a problem discussed in this chapter.

\textbf{Hint:} Consider the implementation details and potential challenges.
\end{problem}

\begin{problem}[Advanced Topic 7]
Evaluate the limitations and potential improvements of techniques from this chapter.

\textbf{Hint:} Consider both theoretical limitations and practical constraints.
\end{problem}
