% Chapter 20: Deep Generative Models

\chapter{Deep Generative Models}
\label{chap:deep-generative-models}

This chapter examines modern approaches to generating new data samples using deep learning.


\section*{Learning Objectives}
\addcontentsline{toc}{section}{Learning Objectives}

After studying this chapter, you will be able to:

\begin{enumerate}
    \item Compare VAEs, GANs, and flow/diffusion models conceptually and practically.
    \item Write training objectives and sampling procedures for each class.
    \item Evaluate generative models with proper metrics and qualitative checks.
    \item Discuss trade-offs in likelihood, sample quality, and mode coverage.
\end{enumerate}



\section*{Intuition}
\addcontentsline{toc}{section}{Intuition}

Generative models are machine learning systems that learn to create new data samples that resemble a given dataset. For example, after training on thousands of cat photos, a generative model can produce entirely new, realistic cat images that never existed before.

Think of generative models as digital artists who study masterpieces to understand artistic style, then create original works in that same style. Just as a painter learns brushstrokes and color palettes from studying great works, these models learn the underlying patterns and structures in data to generate novel samples.

Generative models learn data distributions in complementary ways: explicit likelihoods, implicit adversarial training, or score-based diffusion. Each chooses a tractable learning signal that captures structure while managing complexity.


\input{chapters/chap20-sec01}
\input{chapters/chap20-sec02}
\input{chapters/chap20-sec03}
\input{chapters/chap20-sec04}
\input{chapters/chap20-sec05}

\input{chapters/chap20-real-world-applications}

% Chapter summary and problems
\input{chapters/chap20-key-takeaways}
\input{chapters/chap20-exercises}
