% Exercises (Exercises) for Chapter 15

\section*{Exercises}
\addcontentsline{toc}{section}{Exercises}

\subsection*{Easy}

\begin{problem}[Encoder-Decoder Symmetry]
Why share architecture between $q(z|x)$ and $p(x|z)$?

\textbf{Hint:} Computation; conceptual symmetry.
\end{problem}

\begin{problem}[KL in ELBO]
State the role of $D_{KL}(q||p)$ in VAE training.

\textbf{Hint:} Regularisation; posterior matching.
\end{problem}

\begin{problem}[Reparameterisation Trick]
Explain why reparameterisation enables gradient flow.

\textbf{Hint:} Sampling vs. deterministic path.
\end{problem}

\begin{problem}[Prior Choice]
Justify Gaussian prior for VAE latents.

\textbf{Hint:} Tractability; simplicity; universality.
\end{problem}

\subsection*{Medium}

\begin{problem}[ELBO Derivation]
Derive the ELBO from Jensen's inequality.

\textbf{Hint:} $\log \mathbb{E}[X] \geq \mathbb{E}[\log X]$.
\end{problem}

\begin{problem}[Beta-VAE]
Explain how $\beta$-VAE encourages disentanglement.

\textbf{Hint:} Weighted KL penalty; independence.
\end{problem}

\subsection*{Hard}

\begin{problem}[Posterior Collapse]
Analyse conditions causing posterior collapse and propose mitigation.

\textbf{Hint:} Strong decoder; KL annealing; free bits.
\end{problem}

\begin{problem}[Importance-Weighted ELBO]
Derive the importance-weighted ELBO and show it tightens the bound.

\textbf{Hint:} Multiple samples; log-mean-exp.
\end{problem}


\begin{problem}[Advanced Topic 1]
Explain a key concept from this chapter and its practical applications.

\textbf{Hint:} Consider the theoretical foundations and real-world implications.
\end{problem}

\begin{problem}[Advanced Topic 2]
Analyse the relationship between different techniques covered in this chapter.

\textbf{Hint:} Look for connections and trade-offs between methods.
\end{problem}

\begin{problem}[Advanced Topic 3]
Design an experiment to test a hypothesis related to this chapter's content.

\textbf{Hint:} Consider experimental design, metrics, and potential confounding factors.
\end{problem}

\begin{problem}[Advanced Topic 4]
Compare different approaches to solving a problem from this chapter.

\textbf{Hint:} Consider computational complexity, accuracy, and practical considerations.
\end{problem}

\begin{problem}[Advanced Topic 5]
Derive a mathematical relationship or prove a theorem from this chapter.

\textbf{Hint:} Start with the definitions and work through the logical steps.
\end{problem}

\begin{problem}[Advanced Topic 6]
Implement a practical solution to a problem discussed in this chapter.

\textbf{Hint:} Consider the implementation details and potential challenges.
\end{problem}

\begin{problem}[Advanced Topic 7]
Evaluate the limitations and potential improvements of techniques from this chapter.

\textbf{Hint:} Consider both theoretical limitations and practical constraints.
\end{problem}
