% Chapter 4: Numerical Computation

\chapter{Numerical Computation}
\label{chap:numerical-computation}

This chapter covers numerical methods and computational considerations essential for implementing deep learning algorithms. Topics include gradient-based optimization, numerical stability, and conditioning.

\section*{Learning Objectives}

After studying this chapter, you will be able to:

\begin{itemize}
    \item \textbf{Understand numerical precision issues}: Recognize how finite precision arithmetic affects deep learning computations and implement solutions for overflow, underflow, and numerical instability.
    
    \item \textbf{Master gradient-based optimization}: Apply gradient descent and understand the role of Jacobian and Hessian matrices in optimization landscapes, including critical points and saddle points.
    
    \item \textbf{Handle constrained optimization}: Use Lagrange multipliers and KKT conditions to solve optimization problems with constraints, relevant for regularization and fairness in deep learning.
    
    \item \textbf{Assess numerical stability}: Calculate condition numbers, identify ill-conditioned problems, and implement gradient checking and other numerical verification techniques.
    
    \item \textbf{Apply practical numerical techniques}: Use log-sum-exp tricks, mixed precision training, and other numerical stability methods in real deep learning implementations.
    
    \item \textbf{Debug numerical issues}: Recognize common numerical problems in deep learning and apply appropriate solutions for robust model training.
\end{itemize}

\input{chapters/chap04-sec01}
\input{chapters/chap04-sec02}
\input{chapters/chap04-sec03}
\input{chapters/chap04-sec04}

% Chapter summary and exercises
\input{chapters/chap04-key-takeaways}
\input{chapters/chap04-exercises}
