% Chapter 14: Autoencoders

\chapter{Autoencoders}
\label{chap:autoencoders}

This chapter explores autoencoders, neural networks designed for unsupervised learning through data reconstruction.


\section*{Learning Objectives}
\addcontentsline{toc}{section}{Learning Objectives}

After studying this chapter, you will be able to:

\begin{enumerate}
    \item Describe the autoencoder framework and common variants (denoising, sparse, contractive).
    \item Explain the role of bottlenecks and regularization in learning useful representations.
    \item Implement training objectives and evaluate reconstruction vs. downstream utility.
    \item Understand the connection between autoencoders and generative models.
\end{enumerate}



\section*{Intuition}
\addcontentsline{toc}{section}{Intuition}

Autoencoders are like skilled artists who can recreate a masterpiece from just a few key brushstrokes - they learn to compress complex data into essential features while discarding unnecessary details. The compression process involves reducing the dimensionality of input data while preserving the most important information needed for accurate reconstruction, where the goal is to learn a compact representation that captures the underlying structure and patterns in the data. This compression enables the model to identify and retain only the salient features that are necessary for understanding the data, while discarding noise and redundancies that don't contribute to the essential structure, ultimately surfacing meaningful patterns that can transfer to other tasks and applications.


\input{chapters/chap14-sec01}
\input{chapters/chap14-sec02}
\input{chapters/chap14-sec03}
\input{chapters/chap14-sec04}

\input{chapters/chap14-real-world-applications}

% Chapter summary and problems
\input{chapters/chap14-key-takeaways}
\input{chapters/chap14-exercises}
