% Exercises (Exercises) for Chapter 20

\section*{Exercises}
\addcontentsline{toc}{section}{Exercises}

\subsection*{Easy}

\begin{problem}[VAE vs. GAN]
Compare the training objectives of VAEs and GANs.

\textbf{Hint:} Likelihood-based vs. adversarial.
\end{problem}

\begin{problem}[Normalising Flow Invertibility]
Why must normalising flows be invertible?

\textbf{Hint:} Exact likelihood computation via change of variables.
\end{problem}

\begin{problem}[Diffusion Process]
Describe the forward diffusion process in diffusion models.

\textbf{Hint:} Gradual addition of Gaussian noise.
\end{problem}

\begin{problem}[Sampling Speed]
Compare sampling speed across VAEs, GANs, and diffusion models.

\textbf{Hint:} Single pass vs. iterative refinement.
\end{problem}

\subsection*{Medium}

\begin{problem}[Flow Jacobian]
Derive the change-of-variables formula for a normalising flow.

\textbf{Hint:} $\log p(x) = \log p(z) + \log|\det \frac{\partial f}{\partial z}|$.
\end{problem}

\begin{problem}[Denoising Score Matching]
Explain how diffusion models learn the score function.

\textbf{Hint:} Predict noise; connection to $\nabla_x \log p(x)$.
\end{problem}

\subsection*{Hard}

\begin{problem}[Coupling Layer Design]
Design an invertible coupling layer and prove its properties.

\textbf{Hint:} Affine transformations; partition dimensions.
\end{problem}

\begin{problem}[Guidance Trade-off]
Analyse the trade-off between sample quality and diversity in classifier-free guidance.

\textbf{Hint:} Guidance scale; conditional vs. unconditional scores.
\end{problem}


\begin{problem}[Advanced Topic 1]
Explain a key concept from this chapter and its practical applications.

\textbf{Hint:} Consider the theoretical foundations and real-world implications.
\end{problem}

\begin{problem}[Advanced Topic 2]
Analyse the relationship between different techniques covered in this chapter.

\textbf{Hint:} Look for connections and trade-offs between methods.
\end{problem}

\begin{problem}[Advanced Topic 3]
Design an experiment to test a hypothesis related to this chapter's content.

\textbf{Hint:} Consider experimental design, metrics, and potential confounding factors.
\end{problem}

\begin{problem}[Advanced Topic 4]
Compare different approaches to solving a problem from this chapter.

\textbf{Hint:} Consider computational complexity, accuracy, and practical considerations.
\end{problem}

\begin{problem}[Advanced Topic 5]
Derive a mathematical relationship or prove a theorem from this chapter.

\textbf{Hint:} Start with the definitions and work through the logical steps.
\end{problem}

\begin{problem}[Advanced Topic 6]
Implement a practical solution to a problem discussed in this chapter.

\textbf{Hint:} Consider the implementation details and potential challenges.
\end{problem}

\begin{problem}[Advanced Topic 7]
Evaluate the limitations and potential improvements of techniques from this chapter.

\textbf{Hint:} Consider both theoretical limitations and practical constraints.
\end{problem}
