% Exercises (Exercises) for Chapter 19

\section*{Exercises}
\addcontentsline{toc}{section}{Exercises}

\subsection*{Easy}

\begin{problem}[Why Approximate?]
Explain why exact inference is intractable in many models.

\textbf{Hint:} Partition function; high-dimensional integration.
\end{problem}

\begin{problem}[ELBO Connection]
Relate ELBO to KL divergence between $q$ and $p$.

\textbf{Hint:} $\log p(x) = \text{ELBO} + D_{KL}(q||p)$.
\end{problem}

\begin{problem}[Mean-Field Assumption]
State the mean-field independence assumption.

\textbf{Hint:} Factored variational distribution.
\end{problem}

\begin{problem}[MCMC vs. VI]
Compare MCMC and variational inference trade-offs.

\textbf{Hint:} Asymptotic exactness vs. computational speed.
\end{problem}

\subsection*{Medium}

\begin{problem}[Coordinate Ascent VI]
Derive the coordinate ascent update for a simple model.

\textbf{Hint:} Fix all but one factor; optimise w.r.t. remaining factor.
\end{problem}

\begin{problem}[Importance Sampling]
Explain how importance sampling estimates expectations.

\textbf{Hint:} Reweight samples from proposal distribution.
\end{problem}

\subsection*{Hard}

\begin{problem}[Amortised Inference]
Analyse the trade-offs of amortised inference in VAEs.

\textbf{Hint:} Amortisation gap; scalability.
\end{problem}

\begin{problem}[Reparameterisation Gradients]
Derive the reparameterisation gradient for a Gaussian variational distribution.

\textbf{Hint:} $z = \mu + \sigma \epsilon$ where $\epsilon \sim \mathcal{N}(0,1)$.
\end{problem}


\begin{problem}[Advanced Topic 1]
Explain a key concept from this chapter and its practical applications.

\textbf{Hint:} Consider the theoretical foundations and real-world implications.
\end{problem}

\begin{problem}[Advanced Topic 2]
Analyse the relationship between different techniques covered in this chapter.

\textbf{Hint:} Look for connections and trade-offs between methods.
\end{problem}

\begin{problem}[Advanced Topic 3]
Design an experiment to test a hypothesis related to this chapter's content.

\textbf{Hint:} Consider experimental design, metrics, and potential confounding factors.
\end{problem}

\begin{problem}[Advanced Topic 4]
Compare different approaches to solving a problem from this chapter.

\textbf{Hint:} Consider computational complexity, accuracy, and practical considerations.
\end{problem}

\begin{problem}[Advanced Topic 5]
Derive a mathematical relationship or prove a theorem from this chapter.

\textbf{Hint:} Start with the definitions and work through the logical steps.
\end{problem}

\begin{problem}[Advanced Topic 6]
Implement a practical solution to a problem discussed in this chapter.

\textbf{Hint:} Consider the implementation details and potential challenges.
\end{problem}

\begin{problem}[Advanced Topic 7]
Evaluate the limitations and potential improvements of techniques from this chapter.

\textbf{Hint:} Consider both theoretical limitations and practical constraints.
\end{problem}
