% Chapter 12: Applications

\chapter{Applications}
\label{chap:applications}

This chapter showcases deep learning applications across various domains, demonstrating the breadth and impact of the field.


\begin{learningobjectives}
\objective{Deep learning opportunities in vision, language, speech, and tabular domains}
\objective{Model families and data representations for different application areas}
\objective{Evaluation protocols and metrics for different task types}
\objective{Common failure modes and deployment considerations}
\end{learningobjectives}



\section*{Intuition}
\addcontentsline{toc}{section}{Intuition}

Applications differ not only in architectures but in \emph{data assumptions and metrics}, requiring practitioners to think beyond model selection to understand the fundamental characteristics of each domain. A robust recipe for approaching any application is: clarify the task and loss function, understand the data distribution and constraints, then choose the simplest model that can plausibly meet the requirements before scaling up.

This systematic approach prevents the common pitfall of over-engineering solutions before understanding the problem's core requirements. For example, a computer vision task might require real-time inference on mobile devices, constraining the choice of architectures regardless of theoretical performance, while a natural language processing application might prioritize interpretability over raw accuracy for regulatory compliance. The key insight is that successful applications balance technical sophistication with practical constraints, often achieving better results through careful problem formulation and data understanding than through complex model architectures.

The applications in this chapter demonstrate how different domains require different approaches: computer vision emphasizes spatial relationships and translation invariance, natural language processing focuses on sequential patterns and semantic understanding, while tabular data applications often require careful feature engineering and handling of missing values. Each domain has its own evaluation metrics, failure modes, and deployment considerations that must be understood before selecting appropriate model families and training strategies.

Real-world success depends on recognizing that the "best" model is not always the most complex or highest-performing on benchmark datasets, but rather the one that best fits the specific constraints, requirements, and context of the intended application. This chapter provides the framework for making these critical decisions systematically across diverse application domains.


\input{chapters/chap12-sec01}
\input{chapters/chap12-sec02}
\input{chapters/chap12-sec03}
\input{chapters/chap12-sec04}
\input{chapters/chap12-sec05}
\input{chapters/chap12-sec06}
% \input{chapters/chap12-real-world-applications}
\input{chapters/chap12-sec07}

% Chapter summary and problems
\input{chapters/chap12-key-takeaways}
\input{chapters/chap12-exercises}
