% Exercises (Exercises) for Chapter 12

\section*{Exercises}
\addcontentsline{toc}{section}{Exercises}

\subsection*{Easy}

\begin{problem}[Vision Metrics]
Choose between accuracy, AUROC, and mAP for object detection. Justify.

\textbf{Hint:} Class imbalance and localisation vs. classification.
\end{problem}

\begin{problem}[NLP Tokenisation]
Explain implications of BPE vs. WordPiece for rare words.

\textbf{Hint:} Subword frequency, OOV handling, sequence length.
\end{problem}

\begin{problem}[Tabular Baselines]
Why can tree ensembles outperform deep nets on small tabular datasets?

\textbf{Hint:} Inductive bias, feature interactions, sample efficiency.
\end{problem}

\begin{problem}[Data Privacy]
List two privacy risks when deploying medical models and mitigations.

\textbf{Hint:} Re-identification, membership inference; anonymisation, DP.
\end{problem}

\subsection*{Medium}

\begin{problem}[Dataset Shift]
Design a test to detect covariate shift between training and production.

\textbf{Hint:} Train a domain classifier; compare feature distributions.
\end{problem}

\begin{problem}[Active Learning]
Propose an acquisition strategy for labelling a limited budget.

\textbf{Hint:} Uncertainty sampling, diversity, class balance.
\end{problem}

\subsection*{Hard}

\begin{problem}[Cost-aware Serving]
Formalise an objective trading accuracy vs. compute cost and latency.

\textbf{Hint:} Multi-objective optimisation; constrained maximisation.
\end{problem}

\begin{problem}[Ethical Deployment]
Specify post-deployment monitoring and escalation for high-stakes tasks.

\textbf{Hint:} Thresholds, audits, rollback, human oversight.
\end{problem}



\begin{problem}[Advanced Topic 1]
Explain a key concept from this chapter and its practical applications.

\textbf{Hint:} Consider the theoretical foundations and real-world implications.
\end{problem}

\begin{problem}[Advanced Topic 2]
Analyse the relationship between different techniques covered in this chapter.

\textbf{Hint:} Look for connections and trade-offs between methods.
\end{problem}

\begin{problem}[Advanced Topic 3]
Design an experiment to test a hypothesis related to this chapter's content.

\textbf{Hint:} Consider experimental design, metrics, and potential confounding factors.
\end{problem}

\begin{problem}[Advanced Topic 4]
Compare different approaches to solving a problem from this chapter.

\textbf{Hint:} Consider computational complexity, accuracy, and practical considerations.
\end{problem}

\begin{problem}[Advanced Topic 5]
Derive a mathematical relationship or prove a theorem from this chapter.

\textbf{Hint:} Start with the definitions and work through the logical steps.
\end{problem}

\begin{problem}[Advanced Topic 6]
Implement a practical solution to a problem discussed in this chapter.

\textbf{Hint:} Consider the implementation details and potential challenges.
\end{problem}

\begin{problem}[Advanced Topic 7]
Evaluate the limitations and potential improvements of techniques from this chapter.

\textbf{Hint:} Consider both theoretical limitations and practical constraints.
\end{problem}
